---
title: "Exercise #3"
subtitle: "Fortgeschrittene Statistische Software für NF"
author: "Lirona Iseni (12591525), Name (Martrikelnummer), ..."
date: "`r Sys.Date()`"
output: distill::distill_article
editor_options: 
  markdown: 
    wrap: 72
---

## General Remarks

-   You can submit your solutions in teams of up to 3 students.
-   Include all your team-member's names and student numbers
    (Matrikelnummern) in the `authors` field.
-   Please use the exercise template document to work on and submit your
    results.
-   Use a level 2 heading for each new exercise and answer each subtask
    next to its bullet points or use a new level 3 heading if you want.
-   Always render the R code for your solutions (`echo=TRUE`) and make
    sure to include the resulting data in your rendered document.
    -   Make sure to not print more than 10 rows of data (unless
        specifically instructed to).
-   Always submit both the rendered document(s) as well as your source
    Rmarkdown or Quarto document. Submit the files separately on moodle,
    **not** as a zip archive.
-   Submission format is HTML. Other formats will lead to a deduction of
    points.

## Exercise 1: Initializing git (4 Points)

For this whole exercise sheet we will be tracking all our changes to it
in git.

a)  Start by initializing a new R project with git support, called
    `2025-exeRcise-sheet-3`. If you forgot how to do this, you can
    follow this
    [guide](https://malikaihle.github.io/Introduction-RStudio-Git-GitHub/rstudio_project.html).
b)  Commit the files generated by Rstudio.
c)  For all of the following tasks in this exercise sheet we ask you to
    always commit your changes after finishing each subtask e.g. create
    a commit after task *1d*, *1e* etc.

> Note: This applies only to answers that have text or code as their
> answer. If you complete tasks in a different order or forget to commit
> one, this is no problem. If you change your answers you can just
> create multiple commits to track the changes.

d)  Name 2 strengths and 2 weaknesses of git. (Don't forget to create a
    commit after this answer, see *1c*)

    # Exercise 1d: Git Strengths & Weaknesses

# Strengths:

# 1. Distributed Version Control: Every user has a full repository history, enabling offline work and redundancy.

# 2. Branching/Merging: Isolate features/fixes in branches and merge them seamlessly.

# Weaknesses:

# 1. Learning Curve: Complex commands (e.g., rebase, conflict resolution) challenge beginners.

# 2. Large File Handling: Struggles with binary files (e.g., .xlsx, .png) without Git LFS.

e)  Knit this exercise sheet. Some new files will automatically be
    generated when knitting the sheet e.g. the HTML page. Ignore these
    files, as we only want to track the source files themselves. You
    can, but don't need to create a `.gitignore` file. Just do not
    commit these files manually.

## Exercise 2: Putting your Repository on GitHub (3 Points)

For this task you will upload your solution to GitHub.

a)  Create a new repository on GitHub in your account named
    `exeRcise-sheet-3`. Make sure you create a **public repository** so
    we are able to see it for grading. Add the link to the repository
    below:
b)  Push your code to this new repository by copying and executing the
    snippet on github listed under
    `…or push an existing repository from the command line`.
c)  Regularly push your latest changes to GitHub again and especially do
    so when you are finished with this sheet.
    
GitHub repository link: https://github.com/lironaisn/exeRcise-sheet-3


## Exercise 3: Pixar Films (4 Points)

Download the `pixar_films` and `public_response` datasets from the
GitHub repository and track them in git.

Link:
<https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-03-11>

For small datasets like these adding them to git is not a problem.

```{r}
download.file(
  url = "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-11/pixar_films.csv",
  destfile = "pixar_films.csv"
)

download.file(
  url = "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-11/public_response.csv",
  destfile = "public_response.csv"
)
```


a)  Load the `pixar_films` dataset into R. Clean the dataset by removing
    films without a title. Inspect the variable `film_rating`. What are
    the possible values and what do they mean? Create a factor variable
    for the film rating. Why is this appropriate?
```{r}
library(tidyverse)

pixar_films <- read_csv("pixar_films.csv") %>%
  filter(!is.na(film)) %>%
  mutate(
    film_rating = case_when(
      film_rating == "N/A" ~ NA_character_,
      film_rating == "Not Rated" ~ NA_character_,
      TRUE ~ film_rating
    ),
    film_rating_fct = factor(film_rating, 
                             levels = c("G", "PG"),
                             ordered = TRUE)
  )
```

b)  Inspect the film titles manually. Which films form a film series? A
    film series can be identified by a common word in the titles of the
    films, often in conjunction with a number in the title,
    e.g. "Despicable Me" and "Despicable Me 2". Create a dataframe which displays a list of the different series with the titles of the films and how many films belong to the series. Output the dataframe.
    
```{r}
series_df <- pixar_films %>%
  mutate(
    series = case_when(
      str_detect(film, "Toy Story") ~ "Toy Story",
      str_detect(film, "Cars") ~ "Cars",
      str_detect(film, "Finding|Dory") ~ "Finding Nemo",
      str_detect(film, "Incredibles") ~ "Incredibles",
      str_detect(film, "Monsters") ~ "Monsters Inc",
      TRUE ~ "Standalone"
    )
  ) %>%
  filter(series != "Standalone") %>%
  group_by(series) %>%
  summarise(
    films = list(film),
    num_films = n(),
    .groups = "drop"
  ) %>%
  mutate(films = map_chr(films, ~paste(sort(.x), collapse = ", ")))

series_df %>%
  knitr::kable(col.names = c("Series", "Films", "Number of Films"))
```
    

c)  Load the `public_response` dataframe into R. Convert the
    `cinema_score` variable into a factor while ensuring the factor
    levels are defined in ascending order, from the lowest to the
    highest score. Combine `public_response` with the `pixar_films`
    dataset using an appropriate merge variable.

```{r}
public_response <- read_csv("public_response.csv") %>%
  mutate(
    cinema_score = factor(cinema_score,
                         levels = c("F", "D", "C", "B", "A-", "A", "A+"),
                         ordered = TRUE)
  )

combined_data <- pixar_films %>%
  inner_join(public_response, by = "film")
```


d)  Choose one of the variables representing the public response and
    create a bar plot for the films belonging to a series. Here are the
    details of the plot:

    -   The film series are represented on the x-axis.
    -   Your chosen public response variable is displayed on the y-axis.
    -   Each film in the series is represented as a separate bar. Bars
        are grouped by film under their respective series on the x-axis.
        Order the bars within a series according to the release date of
        the films.
    -   A title and axis labels for context.

    What do you notice when comparing the scores of the films in a
    series? Do you see any patterns?

```{r}
library(tidyverse)
library(ggplot2)

plot_data <- combined_data %>%
  mutate(
    series = case_when(
      str_detect(film, "Toy Story") ~ "Toy Story",
      str_detect(film, "Cars") ~ "Cars",
      str_detect(film, "Finding|Dory") ~ "Finding Nemo",
      str_detect(film, "Incredibles") ~ "Incredibles",
      str_detect(film, "Monsters") ~ "Monsters Inc",
      TRUE ~ "Standalone"
    )
  ) %>%
  filter(series != "Standalone") %>%
  arrange(series, release_date)

ggplot(plot_data, aes(x = series, y = rotten_tomatoes, fill = fct_inorder(film))) +
  geom_col(position = position_dodge(preserve = "single"), width = 0.7) +
  geom_text(aes(label = rotten_tomatoes), 
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3) +
  labs(
    title = "Pixar Film Series: Rotten Tomatoes Scores",
    subtitle = "Comparing critical reception across franchises",
    x = "Film Series",
    y = "Rotten Tomatoes Score (%)",
    fill = "Film Title"
  ) +
  scale_y_continuous(limits = c(0, 105), expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 7)
  )
```

## Exercise 4: Open Analysis (4 points)

This exercise is a bit more open-ended. You can choose any dataset from
[Our World in Data](https://ourworldindata.org/) and analyze it, while
determining the research question yourself.

a)  Go to <https://github.com/owid/owid-datasets/tree/master/datasets>
    and choose a dataset that interests you. You can have a look at
    <https://ourworldindata.org/> to gather some inspiration.
  
b)  Download the dataset and track it in git.

```{r}
download.file(
  url = "https://raw.githubusercontent.com/owid/owid-datasets/master/datasets/Annual%20world%20population%20growth%20rate%20-%20OWID/Annual%20world%20population%20growth%20rate%20-%20OWID.csv",
  destfile = "world_population_growth.csv"
)
```

c)  Put the name / title of the dataset and a link to it below.

-   Dataset Name: Adding Annual world population growth rate - OWID
-   Link: <https://github.com/owid/owid-datasets/tree/master/datasets/Annual%20world%20population%20growth%20rate%20-%20OWID>

d)  Come up with a (research) question you want to answer with the data
    and briefly explain why you believe this is an interesting question
    within one sentence. It should be a question that can be answered
    with the dataset and using R.
    
Topic: Adding Annual world population growth rate - OWID: How has the global population growth rate changed since 1950, and what are the key trends?
    
e)  Use R to answer your chosen question. Please limit your analysis to
    the functions and techniques we have covered so far in the course.
    You are **not expected** to use advanced statistical models or
    external packages which haven't been introduced.
    
```{r}
pop_growth <- read.csv("world_population_growth.csv")

names(pop_growth) <- c("entity", "year", "growth_rate")

pop_since_1950 <- pop_growth[pop_growth$year >= 1950 & pop_growth$entity == "World", ]

max_growth <- max(pop_since_1950$growth_rate, na.rm = TRUE)
min_growth <- min(pop_since_1950$growth_rate, na.rm = TRUE)
current_growth <- tail(pop_since_1950$growth_rate[!is.na(pop_since_1950$growth_rate)], 1)
year_max_growth <- pop_since_1950$year[which.max(pop_since_1950$growth_rate)]

cat("Global population growth rate analysis since 1950:\n")
cat("- Peak growth rate:", round(max_growth, 4), "% in", year_max_growth, "\n")
cat("- Minimum growth rate since 1950:", round(min_growth, 4), "%\n")
cat("- Current growth rate:", round(current_growth, 4), "%\n")
cat("- Overall trend: The growth rate peaked around", year_max_growth, 
    "and has been", ifelse(current_growth < max_growth, "declining", "increasing"), "since then.\n")
```
    
    
f)  Create a meaningful plot / figure with the dataset. Make sure to
    provide a figure caption (via the chunk options / Rmarkdown) and
    correctly label the figure.
    
```{r}
plot(pop_since_1950$year, pop_since_1950$growth_rate, 
     type = "l", 
     col = "blue", 
     lwd = 2,
     main = "Global Population Growth Rate (1950-Present)",
     xlab = "Year",
     ylab = "Growth Rate (%)",
     ylim = c(0.5, 2.5))

abline(h = 0, col = "gray", lty = 2)

points(year_max_growth, max_growth, col = "red", pch = 19)
text(year_max_growth, max_growth, 
     labels = paste("Peak: ", round(max_growth, 2), "%"), 
     pos = 3, col = "red")

text(max(pop_since_1950$year), current_growth, 
     labels = paste("Current: ", round(current_growth, 2), "%"), 
     pos = 3, col = "darkgreen")

lines(lowess(pop_since_1950$year, pop_since_1950$growth_rate), 
      col = "orange", lwd = 2, lty = 2)
legend("topright", legend = "Trend line", col = "orange", lty = 2, lwd = 2)
```


## Final Note

Make sure to push all your commits and changes to GitHub before
submitting the exercise sheet.
